---
title: "Getting Started with PXN"
author: "Zhining Sui"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PXN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 6, fig.height = 4,   
  out.width = "70%",              
  fig.align = "center",
  dpi = 150
)
```

# Overview
`PXN` provides cross-platform normalization and prediction of gene expression using a probabilistic PCA (PPCA) model with platform-specific affine transformations.

Given two platforms (e.g., A and B), PXN learns a shared latent structure (via PPCA) and platform-specific transforms so you can:

  - Train a PXN model on paired multi-platform data,
  - (Optionally) refine with gradient descent,
  - Predict the expression on the one platform for new samples measured on the other platform,
  - (Optionally) integrate multiple pairwise models into one grand model for indirect predictions (e.g., A $\to$ C via B).

> Think of your gene expression data $Y$ like this (rows = genes, columns = samples): for $K$ platforms, we stack $Y$ by platform: `Y_stacked = rbind(Y_platform1, Y_platform2, ..., Y_platformK)`
> $X$ (optional) is a covariate matrix with columns = samples (one column per sample in $Y$).

This vignette walks through the minimal, practical workflow using the packaged simulation data so everyone can run this vignette as-is: expression matrices in `sim_expres` and matching covariates in `sim_covars`.

We will:

  - Load `sim_expres` and `sim_covars`, understand their shapes, and check alignment.
  - Fit a PXN model on dataset AB (two platforms), optionally refine with GD, and predict A$\to$B on a small holdout set.
  - Show how to run 3-platform training with dataset BCD.
  -  Briefly demonstrate model integration across AB, BCD, EF to enable predictions between indirectly connected platforms.

# Installation
```{r installation}
# Development install:
# install.packages("devtools")
# devtools::install_github("zhiningsui/PXN")

library(PXN)
library(ggplot2) # plotting

```

# Load the built-in simulation dataset
The dataset `sim_expres` is a list with three matrices:

  - `AB`: $(2m)\times n$ stacked by rows (top = A, bottom = B)
  - `BCD`: $(3m)\times n$ stacked (B, C, D)
  - `EF`: $(2m)\times n$ stacked (E, F)

Each matrix has rows = genes, columns = samples.

```{r}
data("sim_expres")
data("sim_covars")

str(sim_expres,  max.level = 1)
str(sim_covars,  max.level = 1)

lapply(sim_expres, dim)

# Sanity check: covariates align to expression by sample columns
stopifnot(
  ncol(sim_expres$AB)  == ncol(sim_covars$AB),
  ncol(sim_expres$BCD) == ncol(sim_covars$BCD),
  ncol(sim_expres$EF)  == ncol(sim_covars$EF)
)
```
**Why this matters:** PXN assumes columns of $X$ and $Y$ refer to the same samples in the same order.

> **Runtime tip:** to keep the vignette quick, we’ll subset to a few hundred genes and ~24 samples. You can remove the subsetting lines to run the full data.

We’ll make tiny, reproducible subsets so the vignette finishes fast:
```{r}
m <- 1000 # total number of genes

# AB
m_keep <- 400  # fast demo; remove to use all genes
Y_AB <- sim_expres$AB[c(1:m_keep, (m+1):(m+m_keep)), 1:24, drop = FALSE]
X_AB <- sim_covars$AB[, 1:24, drop = FALSE]

# BCD (use 300 genes to show different m per dataset)
m_keep <- 300  # fast demo; remove to use all genes
Y_BCD <- sim_expres$BCD[c(1:m_keep, (m+1):(m+m_keep), (2*m+1):(2*m+m_keep)), 1:24, drop = FALSE]
X_BCD <- sim_covars$BCD[, 1:24, drop = FALSE]

# EF
m_keep <- 400
Y_EF <- sim_expres$EF[c(1:m_keep, (m+1):(m+m_keep)), 1:24, drop = FALSE]
X_EF <- sim_covars$EF[, 1:24, drop = FALSE]

```

> **Tip:** Keep gene order consistent across platforms and center/scale as your workflow requires (the package functions assume consistent preprocessing across platforms).

# A. Two-platform workflow on AB
## A1. Split stacked rows into A and B
```{r}
m_AB <- nrow(Y_AB) / 2
stopifnot(m_AB == floor(m_AB)) # rows are stacked as rbind(YA, YB)

YA <- Y_AB[1:m_AB, , drop = FALSE]           # source platform (A)
YB <- Y_AB[(m_AB + 1):(2*m_AB), , drop = FALSE]  # target platform (B)

# Optional double-check: same genes in the same order?
stopifnot(identical(rownames(YA), rownames(YB)))
```

## A2. Make a tiny train/test split
We’ll hold out the last 8 samples to evaluate A$\to$B prediction.
```{r}
n <- ncol(Y_AB)
test_idx <- (n-7):n               # last 8 samples as test
train_idx <- setdiff(seq_len(n), test_idx)

Y_train <- Y_AB[, train_idx, drop = FALSE]
Y_test <- Y_AB[, test_idx,  drop = FALSE]
X_train <- X_AB[, train_idx, drop = FALSE]
X_test <- X_AB[, test_idx,  drop = FALSE]

YA_test <- Y_test[1:m_AB, , drop = FALSE]
YB_true <- Y_test[(m_AB+1):(2*m_AB), , drop = FALSE]

```

## A3. Fit $\to$ (optional) refine via GD $\to$ Predict A$\to$B
### Fit a PXN model
Use `InitEst()` to initialize the model. Set `K=2` for 2 platforms and pick a small latent dimension $L$ (we’ll tune it properly below; we use `L=2` here):

```{r}
K <- 2
L <- 2

# 1) initialize
mod0 <- InitEst(X = X_train, Y = Y_train, K = K, L = L)      # initialize
```

### Optional: refine with gradient descent
```{r}
# 2) optional refinement (often improves prediction)
mod <- GDfun(X = X_train, Y = Y_train, K = K, mod0)       
```
If you don’t want refinement, just use `mod0` in the next steps.

### Predict B from A on the test columns:
```{r}
YB_hat <- Prediction(
  Ysource = YA_test,
  X = X_test,        # columns must align with Y_test
  trained.model = mod,
  k.source = 1,             # 1 = A
  k.target = 2              # 2 = B
)
```

**Evaluate with a simple MSE:**
```{r}
mse_AB <- mean((YB_hat - YB_true)^2)
mse_AB

```

**Visual check (random 80 genes):**
```{r, fig.width=6}
set.seed(999)
g <- sample(seq_len(nrow(YB_true)), 80)
dfp <- data.frame(truth = as.vector(YB_true[g, ]),
                  pred  = as.vector(YB_hat[g, ]))

ggplot(dfp, aes(truth, pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(title = "A → B prediction (subset of genes)", x = "B (truth)", y = "B (pred)") +
  theme_minimal(base_size = 12)

```

## A4. Pick the best latent dimension $L$
A practical way is to sweep a small grid (e.g., `L = 0:4`) and pick the $L$ with the best held-out MSE. For larger studies, use k-fold CV.

**How it works:** for each candidate $L$, we run K-fold CV and record A$\to$B MSE; choose the $L$ with the lowest mean MSE.

```{r}
# General CV helper for any stacked dataset (AB, BCD, EF)
cv_select_L <- function(Y, X, K, L_grid = 0:4, Kfold = 5,
                        pair = c(1, 2), seed = 1, refine = TRUE) {
  stopifnot(length(pair) == 2, pair[1] != pair[2],
            nrow(Y) %% K == 0, ncol(Y) == ncol(X))
  set.seed(seed)
  m <- nrow(Y) / K

  # create K roughly equal folds over sample columns
  idx   <- sample.int(ncol(Y))
  folds <- split(idx, rep_len(seq_len(Kfold), length.out = length(idx)))

  stats <- lapply(L_grid, function(L_try) {
    mses <- numeric(length(folds))
    for (i in seq_along(folds)) {
      test  <- folds[[i]]
      train <- setdiff(seq_len(ncol(Y)), test)

      Y_tr <- Y[, train, drop = FALSE]; X_tr <- X[, train, drop = FALSE]
      Y_te <- Y[,  test, drop = FALSE]; X_te <- X[,  test, drop = FALSE]

      mod0 <- InitEst(X = X_tr, Y = Y_tr, K = K, L = L_try)
      mod  <- if (refine) GDfun(X = X_tr, Y = Y_tr, K = K, mod0) else mod0

      rsrc <- ((pair[1] - 1) * m + 1):(pair[1] * m)  # row block for source platform
      rtgt <- ((pair[2] - 1) * m + 1):(pair[2] * m)  # row block for target platform

      Ysrc_te <- Y_te[rsrc, , drop = FALSE]
      Ytgt_te <- Y_te[rtgt, , drop = FALSE]

      Yhat <- Prediction(Ysource = Ysrc_te, X = X_te,
                         trained.model = mod,
                         k.source = pair[1], k.target = pair[2])

      mses[i] <- mean((Yhat - Ytgt_te)^2)
    }
    c(mean = mean(mses), sd = sd(mses))
  })

  data.frame(
    L    = L_grid,
    mean = sapply(stats, `[`, "mean"),
    sd   = sapply(stats, `[`, "sd")
  )
}


```

Example: Run CV on AB

```{r}
L_grid <- 0:5             # small grid for speed (expand if needed)
cv_ab  <- cv_select_L(Y = Y_AB, X = X_AB, K = 2,
                      L_grid = L_grid, Kfold = 5,
                      pair = c(1, 2), seed = 123, refine = TRUE)
cv_ab
best_L_AB <- cv_ab$L[which.min(cv_ab$mean)]
best_L_AB

```
```{r}
ggplot(cv_ab, aes(L, mean)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.12) +
  labs(y = "CV MSE (A→B)", x = "Latent dimension L",
       title = "Select L via K-fold CV (AB)") +
  theme_minimal(base_size = 12)

```

> Use it: refit AB with `L <- best_L_AB` on all columns of Y_AB/X_AB before making your final predictions.

# B. Three-platform example on BCD
The `BCD` matrix stacks B, C, D by rows. Set `K = 3` and fit just like AB.
```{r}
m_BCD <- nrow(Y_BCD) / 3
K3 <- 3

L3 <- 2
mod0_bcd <- InitEst(X = X_BCD, Y = Y_BCD, K = K3, L = L3)
mod_bcd  <- GDfun(X = X_BCD, Y = Y_BCD, K = K3, mod0_bcd)
mod_bcd$dd[1:min(5, length(mod_bcd$dd))]   # peek at leading singular values

```

You can now call `Prediction()` for any ordered pair among \{B, C, D\} by choosing the appropriate `k.source`, `k.target` in \{1,2,3\}.

### Example: predict B $\to$ C on a tiny holdout:
```{r}
n3 <- ncol(Y_BCD)
ti  <- (n3-7):n3
Yi  <- Y_BCD[, ti, drop = FALSE]
Xi  <- X_BCD[, ti, drop = FALSE]

YB_i <- Yi[1:m_BCD, , drop = FALSE]
YC_i <- Yi[(m_BCD+1):(2*m_BCD), , drop = FALSE]

YC_hat <- Prediction(Ysource = YB_i, X = Xi, trained.model = mod_bcd, k.source = 1, k.target = 2)

mean((YC_hat - YC_i)^2)

```

```{r}
# CV here too: B→D 
cv_select_L(Y_BCD, X_BCD, K = 3, pair = c(1,3))
```

# C. Integrating models
## C1. Integrating models across AB, BCD, EF
If you trained separate models for different platform groups (e.g., AB, BCD, EF), you can integrate them into a single grand model to enable predictions across groups (e.g., A $\to$ C even if A and C were never directly paired).

> **Tip:** When integrating, use a common $L$ across pair models (e.g., pick a weighted average of best $L$ per pair and re-fit).

Below we fit small models on the three datasets with the same `L_int` and call `ModIntegrate()`.
```{r}
L_int <- 2

# AB
mod_AB0 <- InitEst(X_AB, Y_AB, K = 2, L = L_int)
mod_AB <- GDfun(X_AB, Y_AB, K = 2, mod_AB0)

# BCD
mod_BCD0 <- InitEst(X_BCD, Y_BCD, K = 3, L = L_int)
mod_BCD <- GDfun(X_BCD, Y_BCD, K = 3, mod_BCD0)

# EF
mod_EF0 <- InitEst(X_EF, Y_EF,K = 2, L = L_int)
mod_EF <- GDfun(X_EF, Y_EF, K = 2, mod_EF0)

mod_list <- list(AB = mod_AB, BCD = mod_BCD, EF = mod_EF)
Xs_list <- list(AB = X_AB, BCD = X_BCD, EF = X_EF)
Ys_list <- list(AB = Y_AB, BCD = Y_BCD, EF = Y_EF)

# PairList specifies the row-block names in each component (example labels):
pair_list <- list(
  AB = c("YA","YB"),
  BCD = c("YB","YC","YD"),
  EF = c("YE","YF")
)

grand <- ModIntegrate(modList = mod_list, Xlist = Xs_list, Ylist = Ys_list, PairList = pair_list)
names(grand)

```

After integration, you can attempt predictions that weren’t directly trained in any single model. The exact indices for platforms inside `grand` depend on your linkage; inspect `grand` (or your `PairList`) to pick `k.source` and `k.target`. 

## C2. Make predictions with the grand model
We’ll define a tiny helper to reference platforms by name.

If your grand object exposes platform names (e.g., `grand$platform_names`), we’ll use those; otherwise we’ll fall back to the names we provided in `pair_list`.
```{r}
# Platform index helper for the integrated model
platform_names <- if (!is.null(grand$platform_names)) {
  grand$platform_names
} else {
  # fallback to the order implied by our example PairList
  c("YA","YB","YC","YD","YE","YF")
}
idx <- function(nm) {
  i <- match(nm, platform_names)
  if (is.na(i)) stop("Platform '", nm, "' not found in grand model. ",
                    "Inspect names with: print(platform_names)")
  i
}

platform_names  # show what we will use

```

**Gene consistency utilities:** Strict, name-based alignment that reorders to the model’s genes and drops/fills missing as requested.
```{r}
pxn_model_genes <- function(mod) {
  g <- rownames(mod$b0)
  if (is.null(g) && !is.null(mod$U)) g <- rownames(mod$U)
  if (is.null(g) && !is.null(mod$sigma2s)) g <- names(mod$sigma2s)
  if (is.null(g)) stop("Trained model has no gene names on b0/U/sigma2s.")
  if (anyDuplicated(g)) stop("Trained model has duplicated gene names.")
  g
}

pxn_align_to_genes <- function(Y, genes, drop_missing = TRUE) {
  if (is.null(rownames(Y))) stop("Y has no rownames; cannot align by genes.")
  if (anyDuplicated(rownames(Y))) stop("Y has duplicated rownames.")
  if (drop_missing) {
    keep <- intersect(genes, rownames(Y))
    if (!length(keep)) stop("No overlapping genes between Y and model.")
    Y2 <- Y[keep, , drop = FALSE]
    ord <- match(genes[genes %in% keep], rownames(Y2))
    Y2 <- Y2[ord, , drop = FALSE]
    Y2
  } else {
    Y2 <- matrix(NA_real_, nrow = length(genes), ncol = ncol(Y),
                 dimnames = list(genes, colnames(Y)))
    keep <- intersect(genes, rownames(Y))
    if (length(keep)) Y2[keep, ] <- Y[keep, , drop = FALSE]
    Y2
  }
}
```


### Example 1: A→B with the grand model (compare to AB model)
Use the same AB test split (`YA_test`, `YB_true`, `X_test`) from section A2.
```{r}
# 1) model gene names
gnames <- pxn_model_genes(grand)
# 2) align both source and ground truth
YA_test_aligned <- pxn_align_to_genes(YA_test, gnames, drop_missing = TRUE)
YB_true_aligned <- pxn_align_to_genes(YB_true, gnames, drop_missing = TRUE)

stopifnot(identical(rownames(YA_test_aligned), rownames(YB_true_aligned)))

# 3) predict with grand
YB_hat_grand <- Prediction(
  Ysource = YA_test_aligned,
  X = X_test,
  trained.model = grand,
  k.source = idx("YA"),
  k.target = idx("YB")
)

mse_AB_grand <- mean((YB_hat_grand - YB_true_aligned)^2)
c(AB_pair_model = mse_AB, AB_grand = mse_AB_grand)

```

### Example 2: B→C with the grand model (compare to BCD model)
Reuse the BCD test split from section B (objects: `YB_i`, `YC_i`, `Xi`) and align to grand.

```{r}
# Reuse the BCD tiny holdout (YB_i, YC_i, Xi) and align to grand
YB_i_aligned <- pxn_align_to_genes(YB_i, gnames, drop_missing = TRUE)
YC_i_aligned <- pxn_align_to_genes(YC_i, gnames, drop_missing = TRUE)

YC_hat_grand <- Prediction(
  Ysource = YB_i_aligned,
  X = Xi,
  trained.model = grand,
  k.source = idx("YB"),
  k.target = idx("YC")
)

mse_BC_pair  <- mean((Prediction(Ysource = YB_i, X = Xi,
                                 trained.model = mod_bcd,
                                 k.source = 1, k.target = 2) - YC_i)^2)
mse_BC_grand <- mean((YC_hat_grand - YC_i_aligned)^2)

c(BC_pair_model = mse_BC_pair, BC_grand = mse_BC_grand)
```

### Example 3: A→C with the grand model (not trained as a single pair)
```{r}
YC_hat_from_A <- Prediction(
  Ysource = YA_test_aligned,
  X = X_test,
  trained.model = grand,
  k.source = idx("YA"),
  k.target = idx("YC")
)
dim(YC_hat_from_A)

```
> Evaluating indirect paths: If you don’t have matched A/C for the same samples, use held-out multi-platform pairs (if available) or judge by downstream performance (e.g., DE power, clinical prediction).

# Session info
```{r}
sessionInfo()

```

